---
layout: post
title:  "Simulating Flip 7"
date:   2026-06-05 16:47:00 -0400
date-modified: last-modified
reference-location: margin
number-sections: true
categories:
  - Side Projects
  - Simulations
draft: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, fig.height = 4, fig.width = 9)
```

```{r}
library(tidyverse)
```

## Game functions

### Initialization functions

```{r}
new_deck <- function() {
  as.character(sample(c(0, rep(1:12, 1:12))))
}
new_turn <- function() {
  list(hand = c(),
       score = 0,
       prob_of_busting = 0)
}
new_player <- function(.player_num) {
  list(player_num = .player_num,
       turns = NULL,
       score = 0,
       
       busted = FALSE,
       busting_card = NA,
       stayed = FALSE,
       staying_reason = NULL,
       
       # strategies
       prob_tol = 0.25,
       min_score = 25,
       max_cards = 4
  )
}
```

### Getters

```{r}
whose_turn_is_it <- function(.round, .turn) {
  
  # Modulo turn num by n_players
  mod <- .turn %% .round$n_players
  
  # The mod is the player number, except when it divides evenly (4 %% 4 = 0, so turn that into player 4)
  player <- if_else(mod == 0, .round$n_players, mod)
  
  return(player)
}
```

```{r}
get_this_turn_num <- function(.round, .player_num) {
  length(.round$players[[.player_num]]$turns)
}
get_this_turn <- function(.round, .player_num) {
  this_turn_num <- get_this_turn_num(.round, .player_num)
  .round$players[[.player_num]]$turns[[this_turn_num]]
}
```


```{r}
is_busted <- function(.round, .player_num) {
  .round$players[[.player_num]]$busted
}
is_stayed <- function(.round, .player_num) {
  .round$players[[.player_num]]$stayed
}
```

Get the number of players left. Used to see if the round is over.

```{r}
n_players_left <- function(.round) {
  busted_stats <- c()
  for (i_player in 1:.round$n_players) {
    busted_stats[i_player] <- .round$players[[i_player]]$busted
  }
  return(sum(busted_stats))
}
```



### Gameplay functions

Deal one card. It takes the top card off the deck, removes it from the deck, and now places it in the "dealt card" status. It's not actually given to a player yet.

```{r}
deal_one_card <- function(.round, .player_num) {
  
  # Get the card
  dealt_card <- .round$deck[1]

  # Remove it from the deck
  .round$deck <- .round$deck[-1]
  
  # That is now the dealt card
  .round$dealt_card <- dealt_card

  # Return the modified round object
  return(.round)
}
```

Set up scaffolding for a turn for a player. This adds to the `$turns` part of their object.

```{r}
set_up_turn <- function(.round, .player_num) {
  
  # Get the number of turns the player has already had.
  n_turns <- length(.round$players[[.player_num]]$turns)
  turn_num <- n_turns + 1
  
  .round$players[[.player_num]]$turns[[turn_num]] <- new_turn()
  
  return(.round)
}
```


```{r}
busted <- function(.round, .player_num) {
  
  this_turn_num <- get_this_turn_num(.round, .player_num)
  
  # This is only relevant after the first turn
  if (this_turn_num > 1) {
    
    # Check to see if the dealt card is in the previous turn's hand.
    # The current turn's hand is empty still because we haven't called `add_to_hand` yet.
    is_duplicate <- .round$dealt_card %in% .round$players[[.player_num]]$turns[[this_turn_num-1]]$hand
    
    # If it's a duplicate, toss their hand and mark them as busted.
    if (is_duplicate) {
      .round$players[[.player_num]]$busted <- TRUE
      .round$players[[.player_num]]$busting_card <- .round$dealt_card
    }
  }
  
  return(.round)
}
```

Player adds the dealt card to their hand. We've already determined if they're busted, so if they are, it's not added. Then remove the card from the "dealt card" status.

```{r}
add_to_hand <- function(.round, .player_num) {
  
  this_turn_num <- get_this_turn_num(.round, .player_num)
  
  # Add it to their hand (if they're not busted)
  if (!is_busted(.round, .player_num)) {
    
    # If it's the first turn, the dealth card becomes the hand.
    if (this_turn_num == 1) {
      .round$players[[.player_num]]$turns[[this_turn_num]]$hand <- .round$dealt_card
      
    # If it's not, then add the dealt card to the previous turn's hand
    } else {
      .round$players[[.player_num]]$turns[[this_turn_num]]$hand <- 
        c(.round$players[[.player_num]]$turns[[this_turn_num-1]]$hand, .round$dealt_card)
    }
  }

  # It's no longer the dealt card
  .round$dealt_card <- NA
  
  return(.round)
}

```


### End-of-turn functions

This will need to be changed a lot, but converting to numbers is fine for now.

```{r}
calculate_scores <- function(.round) {
  for (i_player in 1:.round$n_players) {
    this_turn_num <- get_this_turn_num(.round, i_player)
    
    # Don't do this if a player hasn't had a turn yet.
    if (!is.null(.round$players[[i_player]]$turns[[this_turn_num]])) {
      .round$players[[i_player]]$turns[[this_turn_num]]$score <-
        .round$players[[i_player]]$turns[[this_turn_num]]$hand |> as.numeric() |> sum()
    }
  }
  return(.round)
}
```

```{r}
calculate_probs <- function(.round) {
  
  cards_in_deck <- length(.round$deck)
  
  for (i_player in 1:.round$n_players) {
    this_turn_num <- get_this_turn_num(.round, i_player)
    
    # Don't bother if a player hasn't had a turn yet.
    if (this_turn_num > 0) {
      n_potential_duplicates <- sum(.round$deck %in% .round$players[[i_player]]$turns[[this_turn_num]]$hand)
      .round$players[[i_player]]$turns[[this_turn_num]]$prob_of_busting <- n_potential_duplicates/cards_in_deck
    }
  }
  
  return(.round)
}
```





### Strategy

I have three strategies built in:

* probability tolerance (`prob_tol`): The maximum probability of busting someone is willing to move forward. A default of 1 means they'll always go until busted. A score of 0.25 means they'll stay if the probability of busting exceeds 0.25. 
* minimum score (`min_score`): The lowest score someone is willing to stay with. The maximum (around 67) means they'll go until they bust. A minimum score of 25 means they'll stay as soon as they get 25 points.
* maximum number of cards (`max_cards`): The most number of cards someone is willing to stay with. The maxmimum (of 7) means they'll go until busted. A score of 4 means they stay with just four cards.

```{r}
decide_to_stay <- function(.round, .player_num) {
  this_prob_tol  <- .round$players[[.player_num]]$prob_tol
  this_min_score <- .round$players[[.player_num]]$min_score
  this_max_cards <- .round$players[[.player_num]]$max_cards
  
  this_turn_num <- get_this_turn_num(.round, .player_num)
  
  current_prob    <- .round$players[[.player_num]]$turns[[this_turn_num]]$prob_of_busting
  current_score   <- .round$players[[.player_num]]$turns[[this_turn_num]]$score
  current_n_cards <- length(.round$players[[.player_num]]$turns[[this_turn_num]]$hand)
  
  if (current_prob >= this_prob_tol) {
    .round$players[[.player_num]]$stayed <- TRUE
    msg <- paste0("Current probability (", current_prob, ") exceeds my max tolerable probability (", this_prob_tol, ").")
    .round$players[[.player_num]]$staying_reason <- c(.round$players[[.player_num]]$staying_reason, msg)
  }
  
  if (current_score >= this_min_score) {
    .round$players[[.player_num]]$stayed <- TRUE
    msg <- paste0("Current score (", current_score, ") exceeds the max score I want (", this_min_score, ").")
    .round$players[[.player_num]]$staying_reason <- c(.round$players[[.player_num]]$staying_reason, msg)
  }
  
  if (current_n_cards >= this_max_cards) {
    .round$players[[.player_num]]$stayed <- TRUE
    msg <- paste0("Current number of cards (", current_n_cards, ") exceeds the max number of cards I want (", this_max_cards, ").")
    .round$players[[.player_num]]$staying_reason <- c(.round$players[[.player_num]]$staying_reason, msg)
  }
  
  return(.round)
}
decide_to_stay(round, 3)
```



### The round

```{r}
run_round <- function(.n_players = 4) {
  
  round <- list(
    n_players = .n_players,
    deck = new_deck(),
    players = map(1:.n_players, new_player),
    dealt_card = NA,
    winner = NA
  )
  
  for (i_turn in 1:(.n_players * 7)) {
    whose_turn <- whose_turn_is_it(round, i_turn)

    # Skip this player if they're busted or stayed
    if (is_busted(round, whose_turn)) { next }
    if (is_stayed(round, whose_turn)) { next }

    # Set up scaffolding for this player's turn.
    round <- set_up_turn(round, whose_turn)

    # Deal one card
    round <- deal_one_card(round, whose_turn)

    # Busted if it's a duplicate
    round <- busted(round, whose_turn)

    # Give it to the player
    round <- add_to_hand(round, whose_turn)

    # End of round functions
    round <- calculate_scores(round)
    round <- calculate_probs(round)
    
    # Add the final score for each round to their total score when people stay
    
    # Decide to stay
    round <- decide_to_stay(round, whose_turn)
    
    # If every player has gone but there's only one left.
    if (i_turn > round$n_players && n_players_left(round) == (round$n_players - 1)) {
      break
    }
  }
  
  return(round)
}
round <- run_round()
round
```

### Full game

Now run a full game to 200 points.

```{r}
add_scores_to_total <- function(.round) {
  scores <- c()
  for (i_player in 1:.round$n_players) {
    scores[[i_player]] <- .round$players[[i_player]]$score
  }
  return(scores)
}
add_scores_to_total(round)

new_game <- function(.n_players = 4) {
  game <- list(rounds = c(),
               scores = tibble(players = 1:.n_players,
                               score = 0),
               winner = NA)
  
  round_num <- 0
  while (is.na(game$winner)) {
    round_num <- round_num + 1
    
    game$rounds[[round_num]] <- run_round(.n_players)
    
    
    
    if (round_num >= 50) { game$winner = "yes" }
  }
  
  return(game)
  
  # be sure to cycle through who starts
}
game <- new_game()
```





## Many simulations

```{r}
sims <- tibble(sim = 1:10000) |> 
  rowwise() |> 
  mutate(round = list(run_round())) |> 
  print()
```



```{r}
get_probs <- function(.round) {
  
  probs <- crossing(player = 1:.round$n_players,
                    turn = 1:7,
                    stats = NA)
  
  for (i_player in 1:.round$n_players) {
    n_turns <- get_this_turn_num(.round, i_player)
    for (i_turn in 1:n_turns) {
      probs[probs$player == i_player & probs$turn == i_turn,]$stats <-
        list(.round$players[[i_player]]$turns[[i_turn]])
    }
  }
  probs_tbl <- probs |> 
    unnest_wider(stats) |> 
    rowwise() |> 
    # mutate(hand = str_c(sort(as.numeric(hand)), collapse = ",")) |> 
    
    unnest(hand) |> 
    mutate(hand = factor(hand, levels = as.character(0:12)),
           held = TRUE,
           .after = hand) |>
    pivot_wider(names_from  = hand,
                names_prefix = "has_",
                values_from  = held,
                values_fill  = FALSE) |> 
    
    filter(!is.na(score))
  
  return(list(stats = probs_tbl))
}

get_probs(sims$round[[1]])

probs <- sims |>
  mutate(stats = get_probs(round)) |>
  select(-round) |> 
  unnest(stats) |> 
  mutate(across(matches("has_"), ~replace_na(., FALSE))) |> 
  select(sim, player, turn, score, prob_of_busting, 
         has_0, has_1, has_2, has_3, has_4, has_5, has_6,
         has_7, has_8, has_9, has_10, has_11, has_12) |> 
  print()
```


```{r}
probs |> 
  ggplot(aes(turn, prob_of_busting, group = turn)) + 
  geom_boxplot() + 
  scale_x_continuous(breaks = 1:7) + 
  scale_y_continuous(breaks = seq(0, 1, 0.2), label = scales::label_percent()) + 
  labs(title = "Probability of busting by hand size",
       x = "number of cards in hand",
       y = "probability of busting")
```

```{r}
probs_lm <- lm(prob_of_busting ~ 
                 turn *  
                 (has_0 + has_1 + has_2 + has_3 + has_4 + 
                    has_5 + has_6 + has_7 + has_8 + has_9 + 
                    has_10 + has_11 + has_12), 
               data = probs)
summary(probs_lm)
```
```{r}
process_potential_hand <- function(.cards) {
  
  tibble(turn = length(.cards),
         has_0  = 0  %in% .cards,
         has_1  = 1  %in% .cards,
         has_2  = 2  %in% .cards,
         has_3  = 3  %in% .cards,
         has_4  = 4  %in% .cards,
         has_5  = 5  %in% .cards,
         has_6  = 6  %in% .cards,
         has_7  = 7  %in% .cards,
         has_8  = 8  %in% .cards,
         has_9  = 9  %in% .cards,
         has_10 = 10 %in% .cards,
         has_11 = 11 %in% .cards,
         has_12 = 12 %in% .cards,)
}
process_potential_hand(c(12))

get_odds <- function(.t) {
  predict(probs_lm, newdata = process_potential_hand(.t))
}
```


```{r}
get_odds(c(12, 9))
```



TODO: Add player tendencies
  - always quit after three cards.
  - always quit if odds > 0.5
  - never quit.